Hydras \& Co.

### 2023-11-20 doc/chapter-primrec.tex
Changes to chapter 11; Renaming some symbols in Epsilon.E0

\textbf{Hint:}  You may use as a helper the function which computes the pair \linebreak
$(\texttt{fib}(n+1),\texttt{fib}(n))$. 
Library \href{../theories/html/hydras.Ackermann.cPair.html}{Ackermann.cPair} contains
the definition of the encoding of $\mathbb{N}^2$ into $\mathbb{N}$, and the proofs that 
the associated constructor and projections are primitive recursive.  
\emph{You may start this exercise with the file
  \href{https://github.com/coq-community/hydra-battles/blob/master/exercises/primrec/FibonacciPR.v}{exercises/primrec/FibonacciPR.v}.}

@@

\section{Proving that a given function is \emph{not} primitive recursive}
\label{sect:ack-not-PR}

The best known example of a total recursive function which is not primitive recursive is the Ackermann function. We show how to adapt the classic proof (see for instance~\cite{planetmath}) to the constraints of \gallina. We hope this formal proof 
 is a nice opportunity to explore
the treatment of primitive recursive functions by R. O'Connor,
and to play with dependent types.

With respect to proof methods, we will see a first \emph{proof by induction on the set of all primitive recursive functions}.
First, we will prove a property shared by all primitive function, 
then show that the Ackermann function doesn't satisfy this property.

@@

\subsection{A proof by induction on all primitive recursive functions}

@@

First, we prove that any primitive function of two arguments is majorized by \texttt{Ack}.
If we look at the inductive definition of primitive recursive functions, page~\pageref{def:Primrec}, it is obvious that a proof by induction on the construction of primitive recursive functions must consider functions of any arity.

The following scheme allows us to write proofs by induction on the class of primitive recursive functions.

@@

First, we extend \texttt{max} to vectors of natural numbers (using the notations of module \texttt{VectorNotations} and some more definitions from 
\href{../theories/html/hydras.Prelude.MoreVectors.html}{Prelude.MoreVectors}). So, (\texttt{t\,$A$\,$n$}) is the type of vectors of $n$ elements of type $A$, and the constants \texttt{cons}, \texttt{nil}, \texttt{map}, etc., refer to vectors and not to lists. Likewise, the notation \texttt{x::v} is an abbreviation for
\texttt{VectorDef.cons x \_ v}.

@@

The remaining cases are proved within a mutual  induction.

@@

Then, we write  a short proof by contradiction.

>>>>

\textbf{Hint:}  You may use as a helper the function which computes the pair \linebreak
$(\texttt{fib}(n+1),\texttt{fib}(n))$. 
Library \href{../theories/html/hydras.Ackermann.cPair.html}{Ackermann.cPair} contains
the definition of the encoding of $\mathbb{N}^2$ into $\mathbb{N}$, and the proofs that 
the associated constructor and projections are primitive recursive.

Please find here some definitions and lemmas you may use 
in order to solve this exercise (non-exhaustive list).

\inputsnippets{FibonacciPR/AboutCPair}

\emph{You may start this exercise with the file
  \href{https://github.com/coq-community/hydra-battles/blob/master/exercises/primrec/FibonacciPR.v}{exercises/primrec/FibonacciPR.v}.}

@@

\section{Proofs by induction over all primitive recursive functions}

We may want to prove properties shared with all the set of
primitive function, for instance that any primitive recursive function is representable by a predicate in Peano Arithmetics (see Chap~›\vref{chap:PRRep}).
In this chapter, we will prove by induction that every primitive recursive function has a given property which is not satisfied by  Ackermann function, and deduce that this function is not PR.
We show how to adapt the classic proof (see for instance~\cite{planetmath}) to the constraints of \gallina. We hope this formal proof 
 is a nice opportunity to explore
the treatment of primitive recursive functions by R. O'Connor,
and to play with dependent types.

First, let us define Ackermann function in \coq, then look at a few properties of this function.

@@

@@

First, we have to prove that any primitive function of two arguments is majorized by \texttt{Ack}.
Let us look at the induction principles generated for the types
\texttt{PrimRec $n$}. % and \texttt{PrimRecs $n$ $m$}

If we look at the inductive definition of primitive recursive functions, page~\pageref{def:Primrec}, it is obvious that a proof by induction on the construction of primitive recursive functions must consider functions of any arity.

@@

First, we extend \texttt{max} to vectors of natural numbers (using the notations of module \texttt{VectorNotations} and some more definitions from 
\href{../theories/html/hydras.Prelude.MoreVectors.html}{Prelude.MoreVectors}). So, (\texttt{t $A$ $n$}) is the type of vectors of $n$ elements of type $A$, and the constants \texttt{cons}, \texttt{nil}, \texttt{map}, etc., refer to vectors and not to lists. Likewise, the notation \texttt{$x$::$v$} is an abbreviation for
\texttt{VectorDef.cons $x$ \_ $v$}.

@@

The remaining cases are proved within the main  mutual  induction.

@@

Then, we write  a short proof by contradiction, using a 
\emph{diagonalized} version of Ackermann function.
